The Importance of Being Nonplanar: Street Network Representation in Urban Form Studies

### 2018-01-14 manuscript.tex
update intro, results

In urban planning and urban design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity. Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they can be used to usefully model pedestrian volume, household travel patterns, urban design patterns and spatial morphology, access and equity, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.

This article presents two new measures of the degree of nonplanarity for urban street networks that can be generalized to any spatial network. It compares the results and interpretation of these nonplanarity indicators to the formal, mathematical planarity of street networks around the world. Despite common claims in the literature that street networks are planar graphs, this study finds that they generally are not and that planar graphs poorly model the street networks of many cities. Further, the magnitude of this bias varies substantially across cities and urbanization types.

This article is organized as follows. The following section introduces the relevant basics of graph theory urban studies, focusing on discussions in the research literature about street network planarity. The next section introduces the methods used to acquire and analyze the street networks in this study. Then this article presents the results of this analysis before concluding with a discussion of their ramifications for street network research and urban form studies.

\section{Background}

Graph theory is the mathematical study of networks \citep{newman_networks:_2010}. Graphs can mathematically model real-world networks such as friendships or the Internet as well as spatial networks, such urban street networks \citep{barthelemy_spatial_2011}. A graph $G$ consists of a set of nodes $N$ connected to one another by a set of edges $E$. An edge $uv$ in a directed graph points in one direction from some node $u$ to some node $v$, but an undirected graph's edges all mutually point in both directions. In a street network, the nodes represent intersections and dead-ends, and the (directed) edges represent the street segments that connect them. How a graph's nodes and edges connect to one another defines its \emph{topology}. For example, a node's \emph{degree} is a topological trait that represents how many edges connect to that node. A \emph{planar} graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes. If the graph \emph{cannot} be drawn or redrawn to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}. Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, and area --- alongside their topology.

@@

This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we shall refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we shall refer to as \emph{spatial planarity}. A street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real-world), but it could still be formally planar. For example, if we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but nodes \citep[for a more comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}. In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.

Consider two theoretical examples: a medieval European city center and a modern American city center. The former's circulation network comprises a set of pedestrian paths and drivable streets without any grade separation, overpasses, or underpasses. In two-dimensions, its edges (i.e., streets and paths) cross each other only at nodes (i.e., intersections), so it is by definition planar. The latter's circulation network comprises pedestrian paths, drivable surface streets, and a grade-separated freeway with overpasses over some of the surface streets. In two-dimensions, its edges occasionally cross each other at non-nodes (i.e., overpasses), so it is by definition nonplanar. However, we might call it \emph{approximately} planar because its nonplanar edge crossings are relatively uncommon. Approximate planarity constrains this nonplanar spatial network such that it does not exhibit certain characteristics found among nonplanar aspatial graphs (e.g., small-world effects or power-law distributed node degrees).

@@

If a planar graph models a street network poorly, it could do so in multiple ways. Forcing planarity on a nonplanar street network creates artificial nodes at bridges and tunnels, which breaks routing. As \citet[p.~6]{kwan_review_1996} puts it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g. recommending that a traveler make a left-turn at an intersection that proves to be an overpass)}. Second, and accordingly, intersection counts and density will be overestimated. Similarly, edge lengths will be underestimated due to these artificial breakpoints. Finally, this bias would likely behave inconsistently across different kinds of cities and street network types.

Given these issues, what do \emph{approximately planar} and \emph{well-modeled} mean for street networks? How close is close enough for a planar graph to competently model a mathematically nonplanar street network? Do the biases of planar models remain consistent across geographies or do they misrepresent different places to different degrees? If street networks are in general approximately planar, then how can we measure how planar or nonplanar a given street network is?

The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, such as its crossing number --- the minimum number of edge crossings of any drawing of $G$ --- and the skewness --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,szekely_successful_2004,chimani_non-planar_2009}. However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}. In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.

@@

Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide. This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents. For cities, such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center. We look separately at their drivable and their walkable street networks.

@@

OpenStreetMap graphs contain many interstitial nodes in the middle of street segments (forming an expansion graph) to allow streets to curve through space via a series of straight-line approximations. OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry. This provides us with an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of our networks.

@@

The first is a formal test of aspatial planarity using the algorithm described by \citet{boyer_subgraph_2012}. That is, is it possible to rearrange the graph's nodes and edges in space --- while preserving its topology --- so that edges cross only at nodes? This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding. However, street networks are spatially embedded, so we develop the following two continuous measures to assess the \enquote{extent} to which they are planar.

The second measure is the spatial planarity ratio (SPR). It represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph). It indicates the extent to which planarity overstates intersections and connectivity in each street network. A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values suggest the extent to which the network is nonplanar.

The third measure is the edge length ratio (ELR). It represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph. It indicates the extent to which planarity understates edge lengths in each street network. A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values suggest the extent to which the network is nonplanar.

Finally, we explore how these indicators might vary across a single municipality. To do so we analyze the drivable street network of Oakland, California, a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown. First we analyze the entire city of Oakland. Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street network centered on each. The resulting statistical dispersion of planarity indicators demonstrates the extent to which agglomerating an entire city's neighborhoods into a single graph may obscure neighborhood-scale effects and characteristics.

@@

@@

Mogadishu is the only city to show perfect planarity across all three indicators and both network types. All three Italian cities show perfect planarity in their drivable networks, but not in their walkable networks.

@@

\begin{figure*}[htbp]
    \center
    \includegraphics[width=0.5\textwidth]{regression_phi_split.png}
    \caption{Log-log plot of ELR vs SPR with simple regression lines: drive $r^2=0.98$ and walk $r^2=0.99$.}
    \label{fig:regression_split}
\end{figure*}

>>>>

In urban planning and design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity. Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.

In the urban studies literature, street networks are typically referred to as \enquote{planar} or \enquote{approximately planar} (see Table \ref{tab:planar_quotes}), meaning that they can be well-modeled by a flat two-dimensional model that inherently forbids overpasses or underpasses. Of course in the real-world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels. This leads to two intertwined questions underexplored in the literature today. First, how well do planar graphs model urban street networks? Second, what does the extent to which a network is nonplanar tell us about the city's infrastructure and development?

This study tests the planarity of street networks around the world. It also presents two new measures of the degree of nonplanarity that can be generalized to other types of spatial networks. These indicators help describe the nature of the urban form and transportation infrastructure. Despite common claims in the literature that street networks are planar graphs, this study finds that they generally are nonplanar and that planar graphs poorly model the street networks of many cities. Further, the magnitude of this bias varies substantially across cities and urbanization types.

This article is organized as follows. The following section introduces the relevant basics of graph theory in urban studies, focusing on discussions in the research literature about street network planarity. The next section discusses the methods used to acquire and analyze the street networks in this study. Then we present the results of this analysis before concluding with a discussion of their ramifications for street network research and urban form studies.

\section{Background and Motivation}

Graph theory is the mathematical study of networks \citep{newman_networks:_2010}. Graphs can mathematically model real-world networks such as friendships or the Internet as well as spatial networks, such urban street networks \citep{barthelemy_spatial_2011}. A graph $G$ consists of a set of nodes $N$ connected to one another by a set of edges $E$. An edge $uv$ in a directed graph points in one direction from some node $u$ to some node $v$, but an undirected graph's edges all mutually point in both directions. In a street network, the nodes represent intersections and dead-ends, and the (directed) edges represent the street segments that connect them. How a graph's nodes and edges connect to one another defines its \emph{topology}. For example, a node's \emph{degree} is a topological trait that represents how many edges connect to that node. A \emph{planar} graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes. If the graph \emph{cannot} be drawn --- or redrawn --- to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}. Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, and areas --- along with their topology.

@@

This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we refer to as \emph{spatial planarity}. A street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real-world), but it could still be formally planar. For example, if we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}. In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.

Consider two theoretical examples: a medieval European city center and a modern American city center. The former's circulation network comprises a set of pedestrian paths and drivable streets without any grade separation, overpasses, or underpasses. In two-dimensions, its edges (i.e., streets and paths) cross each other only at nodes (i.e., intersections), so it is by definition planar. The latter's circulation network comprises pedestrian paths, drivable surface streets, and a grade-separated freeway with overpasses over some of the surface streets. In two-dimensions, its edges occasionally cross each other at non-nodes (i.e., overpasses), so it is by definition nonplanar. However, we might call it \emph{approximately} planar because its nonplanar edge crossings are relatively uncommon. Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs (e.g., small-world effects or power-law distributed node degrees).

@@

If a planar graph models a street network poorly, it could do so in multiple ways. Forcing planarity on a nonplanar street network creates artificial nodes at bridges and tunnels, which breaks routing. As \citet[p.~6]{kwan_review_1996} put it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g. recommending that a traveler make a left-turn at an intersection that proves to be an overpass)}. Second, and accordingly, intersection counts and density will be overestimated by the presence of these pseudonodes. Consequently, edge lengths will be underestimated due to these artificial breakpoints in the middle of street segments. Finally, this bias would likely behave inconsistently across different kinds of cities and street network types.

Given these issues, what do \emph{approximately planar} and \emph{well-modeled} mean for street network research? How close is \enquote{close enough} for a planar graph to competently model a formally nonplanar street network? Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different degrees? And if street networks are at least generally approximately planar, then how can we measure just how planar or nonplanar a given street network is?

The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,szekely_successful_2004,chimani_non-planar_2009}. However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}. In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.

@@

Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide. This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents. For cities, such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center. We look separately at each city's drivable and walkable street networks.

@@

OpenStreetMap's raw data contain many interstitial nodes in the middle of street segments (forming an expansion graph) to allow streets to curve through space via a series of straight-line approximations. OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry. This provides us with an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of our networks.

@@

The first is an aspatial test of formal planarity using the algorithm described by \citet{boyer_subgraph_2012}. That is, is it possible to rearrange the graph's nodes and edges in space --- while preserving its topology --- so that edges cross only at nodes? This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding. However, street networks are spatially embedded, so we develop the following two continuous indicators to assess the \enquote{extent} to which they are planar.

The second measure is the spatial planarity ratio (SPR). It represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph). It indicates the extent to which planarity overstates intersections and connectivity in each street network. A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values indicate the extent to which the network is nonplanar.

The third measure is the edge length ratio (ELR). It represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph. It indicates the extent to which planarity understates edge lengths in each street network. A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values indicate the extent to which the network is nonplanar.

Finally, we explore how these indicators might vary across a single municipality. To do so we analyze the drivable street network of Oakland, California, a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown. First we analyze the entire city of Oakland. Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street networks centered on each. The resulting statistical dispersion of planarity indicators demonstrates the extent to which agglomerating an entire city's neighborhoods into a single graph may obscure neighborhood-scale effects and characteristics.

@@

Table \ref{tab:world_cities} presents the results of the analysis of the 50 cities worldwide.

@@

@@

Figure \ref{fig:world_map_bw} shows the distribution of SPR values around the world. While nearly every European city is in the highest subset, indicating their networks are more planar, most American cities are in the lowest subset, indicating their networks are less planar.

Mogadishu is the only city studied that shows perfect planarity across all three indicators and both network types. All three Italian cities show perfect planarity in their drivable networks, but not in their walkable networks.

\begin{figure}[htbp]
	\includegraphics[width=0.5\textwidth]{regression_phi_split.png}
	\caption{Log-log plot of ELR vs SPR with simple regression lines.}
	\label{fig:regression_split}
\end{figure}

Figure \ref{fig:regression_split} shows the relationship between the SPR and ELR indicators across all 50 cities for both network types, as a log-log plot. The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
